### 2017-07-25 sec-related-work.tex
Grammar fixes from Grammarly analysis.

Our model is based heavily upon the prior art of Hidden Markov
Models~\cite{Rabiner:1990:RSR} for sequence labeling tasks. These types of
models are widely applicable and have been used for tasks such as speech
recognition~\cite{Huang:1990:HMM}, part-of-speech
tagging~\cite{Jurafsky:2009:SLP}, and econometrics~\cite{Hamilton:1990:JoE}
and are a member of the more general family of probabilistic graphical
models~\cite{Koller:2009:PGM}.  A major challenge in applying HMMs and
other graphical models successfully to solve a problem is to design an
appropriate architecture of the model, which always varies according to
specific applications.

For example, in part-of-speech tagging~\cite{Jurafsky:2009:SLP} the output
distributions are categorical (distributions over words from a fixed
vocabulary) and the latent states represent the part-of-speech category for
a word. In speech recognition~\cite{Huang:1990:HMM} the output
distributions might be mixtures of Gaussians to predict real-valued vectors
extracted from short windows of a speech signal. In the domain of
econometrics, \citet{Hamilton:1990:JoE} explores HMMs in the context of
``regime-switching.'' In this framing, the goal is to understand how
econometric data changes by modeling discrete changes in ``regime'' as
having an impact on the resulting real-valued vector data observed. The
``regimes'' are represented with some sort of model that can produce
real-valued vector data, such as a multivariate Gaussian or an
auto-regressive model. The analogy with HMMs is that a ``regime'' is a
latent state, and the characterization of the regime itself is the output
distribution for that latent state. Our model can be seen as such a
``regime-switching'' model where the output of the ``regimes'' that
students are switching between are discrete-valued \emph{sequences} (as
opposed to real numbers, vectors of real numbers, or categorical symbols)
and the model used to represent a specific ``regime'' is an (observable)
Markov chain over the observed student actions. We view the switching
between ``regimes'' as the first ``layer'' of our model, and the
transitioning behavior \emph{within} a ``regime'' between the actions a
student takes as the second ``layer'' of our model.

A multi-layered approach to HMM modeling of sequence data has been
performed before in other domains. \citet{Zhang:2004:CVPR}, for example,
explored a two-layer HMM framework for modeling actions in meetings, but
their definition of ``two-layer'' differs from ours. In their formulation,
the ``lower-layer'' level is used to label audio-video action sequences
into basic events, and the ``upper-layer'' is used to label the output of
the lower-layer to discover higher-level office behavior abstractions.
\citet{Oliver:2004:CVIU} propose a similar layered HMM approach
for modeling office activity at multiple different levels of time
granularity. In their approach, each layer $L$ is represented as a ``bank''
of $K$ different HMMs that model sequences of some length $T_L$. At the
bottom layer ($L = 1$), the bank of $K$ HMMs corresponding to that layer is
run on some initial observation data, considering windows of observations
of length $T_1$. Then an output is generated by using the inferential
results of these $K$ HMMs to make a prediction: which of the $K$ HMMs was
most likely to produce that sequence of observations? This output is then
fed to the next layer of HMMs, which considers sequences of length $T_2$
and outputs prediction results as to which of the $K$ HMMs at layer two
were mostly likely to produce the sequence of outputs produced by the
previous layer, and so on.

Our formulation differs from both \citet{Zhang:2004:CVPR} and
\citet{Oliver:2004:CVIU} in that we do not feed the labeled sequence of the
lower level into the input of the higher level. Instead, our lower level is
actually treated using a non-hidden Markov model, and the higher level is
modeling transitions between the $K$ different non-hidden Markov models we
consider. The problem to be solved is similar in that we wish to predict a
``label'' for a sequence of actions a student takes as well as understand
the transition behavior between those labels.  However, one of the
consequences of modeling the lower layer using a non-hidden Markov model
instead of an HMM directly is that the meaning of the $K$ different latent
states can be more clearly captured by using our proposed behavior
representation. If we were to use an HMM as our first layer, the behavior
patterns (like from Figure~\ref{fig:motivating-example}) would instead have
nodes that represent another set of latent states instead of being concrete
actions themselves. Each of these latent states would then be associated
with some other output distribution over the possible concrete actions to
be considered. In order to understand a single behavior pattern uncovered,
one would first have to understand the different output distributions for
the latent states in that pattern in order to understand the meaning of
that latent state. This further complicates the understanding of higher
level patterns because understanding the higher layer patterns requires
understanding the lower layer patterns. By taking a more restrictive view
on the first layer, we are able to produce a representation that can be
more readily interpreted due to the states in our first layer
representation having an immediately clear meaning (the concrete action
they represent).

The 2L-HMM model we propose is more closely related to the Hierarchical
Hidden Markov Model (HHMM) detailed in \citet{Fine:1998:ML}. Here, the
``layers'' are modeled by having the hidden Markov model have two kinds of
transitions.  Horizontal transitions move between states within a layer,
where vertical transitions move between different layers. At the bottom
layer lie the ``production'' states, which output symbols according to some
probability distribution. Our specific model in this case can be modeled as
a HHMM where the horizontal transitions between nodes at the highest layer
(including self-loops) \emph{must} be immediately followed by a vertical
transition to the lower layer. The output probability distributions over
symbols in the lower layer ``production'' level are forced to emit only one
kind of symbol, and vertical transitions are only allowed into the original
higher-layer state that transitioned down into the lower-layer.

@@

HMMs or similar ideas have been previously applied to model education
data~\cite{Shih:2010:EDM,Kizilcec:2013:LAK,Davis:2016:EDM}, but the
previous models are not well tuned toward the student behavior task and
thus cannot adequately address all the aspects of complexity of student
learning behavior.  A main technical contribution of this paper is to
propose a more general HMM that can better adapt to the variations of
student behavior via its variable resolution and nested HMM structure, and
thus enable discovery of more sophisticated behavior patterns and provide
more detailed characterization of student behavior than the previous
models.

For example, \citet{Kizilcec:2013:LAK} assigned students to states
following a rule-based approach based upon when the student submitted the
assignment for a particular week in the course. They investigated how
students transitioned between these states as the course progressed, and
used the sequence of states a student exhibited as a representation for
performing $k$-means clustering of students into related groups. This
differs from our method substantially: we assign students to states using a
probabilistic framework to account for uncertainty in this state
assignment, and jointly learn \emph{representations} for these states,
which are treated as being \emph{latent} as opposed to pre-defined using
some rule (or set of rules).  Furthermore, our model provides more
flexibility in how the time segments are defined, allowing for both finer
(for example, day-by-day) or coarser (for example, month-by-month)
granularity. \citet{Shih:2010:EDM} investigated the use of HMM-based
clustering techniques for automatic discovery of student learning
strategies when solving a particular problem.  This is similar to our
approach in that the description of behavior profiles is a Markov model,
but cannot further characterize each latent state with another informative
HMM. Thus their work can be regarded to modeling ``micro'' behavior,
whereas our model can model both ``micro'' and ``macro'' behavior.

>>>>

Our model is based heavily on the prior art of Hidden Markov
Models~\cite{Rabiner:1990:RSR} for sequence labeling tasks. These types of
models are widely applicable and have been used for tasks such as speech
recognition~\cite{Huang:1990:HMM}, part-of-speech
tagging~\cite{Jurafsky:2009:SLP}, and econometrics~\cite{Hamilton:1990:JoE}
and are a member of the more general family of probabilistic graphical
models~\cite{Koller:2009:PGM}.  A major challenge in applying HMMs and
other graphical models successfully to solve a problem is to design an
appropriate architecture of the model, which always varies according to
specific applications.

For example, in part-of-speech tagging~\cite{Jurafsky:2009:SLP} the output
distributions are categorical (distributions over words from a fixed
vocabulary) and the latent states represent the part-of-speech category for
a word. In speech recognition~\cite{Huang:1990:HMM}, the output
distributions might be mixtures of Gaussians to predict real-valued vectors
extracted from short windows of a speech signal. In the domain of
econometrics, \citet{Hamilton:1990:JoE} explores HMMs in the context of
``regime-switching.'' In this framing, the goal is to understand how
econometric data changes by modeling discrete changes in ``regime'' as
having an impact on the resulting real-valued vector data observed. The
``regimes'' are represented with some sort of model that can produce
real-valued vector data, such as a multivariate Gaussian or an
auto-regressive model. The analogy with HMMs is that a ``regime'' is a
latent state, and the characterization of the regime itself is the output
distribution for that latent state. Our model can be seen as such a
``regime-switching'' model where the output of the ``regimes'' that
students are switching between are discrete-valued \emph{sequences} (as
opposed to real numbers, vectors of real numbers, or categorical symbols)
and the model used to represent a specific ``regime'' is an (observable)
Markov chain over the observed student actions. We view the switching
between ``regimes'' as the first ``layer'' of our model, and the
transitioning behavior \emph{within} a ``regime'' between the actions a
student takes as the second ``layer'' of our model.

A multi-layered approach to HMM modeling of sequence data has been
performed before in other domains. \citet{Zhang:2004:CVPR}, for example,
explored a two-layer HMM framework for modeling actions in meetings, but
their definition of ``two-layer'' differs from ours. In their formulation,
the ``lower-layer'' level is used to label audio-video action sequences
into basic events, and the ``upper-layer'' is used to label the output of
the lower-layer to discover higher-level office behavior abstractions.
\citet{Oliver:2004:CVIU} propose a similar layered HMM approach
for modeling office activity at multiple different levels of time
granularity. In their approach, each layer $L$ is represented as a ``bank''
of $K$ different HMMs that model sequences of some length $T_L$. At the
bottom layer ($L = 1$), the bank of $K$ HMMs corresponding to that layer is
run on some initial observation data, considering windows of observations
of length $T_1$. Then an output is generated by using the inferential
results of these $K$ HMMs to make a prediction: which of the $K$ HMMs was
most likely to produce that sequence of observations? This output is then
fed to the next layer of HMMs, which considers sequences of length $T_2$
and outputs prediction results as to which of the $K$ HMMs at layer two
were most likely to produce the sequence of outputs produced by the
previous layer, and so on.

Our formulation differs from both \citet{Zhang:2004:CVPR} and
\citet{Oliver:2004:CVIU} in that we do not feed the labeled sequence of the
lower level into the input of the higher level. Instead, our lower level is
actually treated using a non-hidden Markov model, and the higher level is
modeling transitions between the $K$ different non-hidden Markov models we
consider. The problem to be solved is similar in that we wish to predict a
``label'' for a sequence of actions a student takes as well as understand
the transition behavior between those labels.  However, one of the
consequences of modeling the lower layer using a non-hidden Markov model
instead of an HMM directly is that the meaning of the $K$ different latent
states can be more clearly captured by using our proposed behavior
representation. If we were to use an HMM as our first layer, the behavior
patterns (like from Figure~\ref{fig:motivating-example}) would instead have
nodes that represent another set of latent states instead of being concrete
actions themselves. Each of these latent states would then be associated
with some other output distribution over the possible concrete actions to
be considered. In order to understand a single behavior pattern uncovered,
one would first have to understand the different output distributions for
the latent states in that pattern in order to understand the meaning of
that latent state. This further complicates the understanding of higher
level patterns because understanding the higher layer patterns requires
understanding the lower layer patterns. By taking a more restrictive view
of the first layer, we are able to produce a representation that can be
more readily interpreted due to the states in our first layer
representation having an immediately clear meaning (the concrete action
they represent).

The 2L-HMM model we propose is more closely related to the Hierarchical
Hidden Markov Model (HHMM) detailed in \citet{Fine:1998:ML}. Here, the
``layers'' are modeled by having the hidden Markov model have two kinds of
transitions.  Horizontal transitions move between states within a layer,
where vertical transitions move between different layers. At the bottom
layer lie the ``production'' states, which output symbols according to some
probability distribution. Our specific model, in this case, can be modeled as
an HHMM where the horizontal transitions between nodes at the highest layer
(including self-loops) \emph{must} be immediately followed by a vertical
transition to the lower layer. The output probability distributions over
symbols in the lower layer ``production'' level are forced to emit only one
kind of symbol, and vertical transitions are only allowed into the original
higher-layer state that transitioned down into the lower-layer.

@@

HMMs or similar ideas have been previously applied to model education
data~\cite{Shih:2010:EDM,Kizilcec:2013:LAK,Davis:2016:EDM}, but the
previous models are not well tuned toward the student behavior task and
thus cannot adequately address all the aspects of the complexity of student
learning behavior.  A main technical contribution of this paper is to
propose a more general HMM that can better adapt to the variations of
student behavior via its variable resolution and nested HMM structure, and
thus enable discovery of more sophisticated behavior patterns and provide
a more detailed characterization of student behavior than the previous
models.

For example, \citet{Kizilcec:2013:LAK} assigned students to states
following a rule-based approach based upon when the student submitted the
assignment for a particular week in the course. They investigated how
students transitioned between these states as the course progressed, and
used the sequence of states a student exhibited as a representation for
performing $k$-means clustering of students into related groups. This
differs from our method substantially: we assign students to states using a
probabilistic framework to account for uncertainty in this state
assignment and jointly learn \emph{representations} for these states,
which are treated as being \emph{latent} as opposed to pre-defined using
some rule (or set of rules).  Furthermore, our model provides more
flexibility in how the time segments are defined, allowing for both finer
(for example, day-by-day) or coarser (for example, month-by-month)
granularity. \citet{Shih:2010:EDM} investigated the use of HMM-based
clustering techniques for automatic discovery of student learning
strategies when solving a particular problem.  This is similar to our
approach in that the description of behavior profiles is a Markov model,
but cannot further characterize each latent state with another informative
HMM. Thus their work can be regarded to modeling ``micro'' behavior,
whereas our model can model both ``micro'' and ``macro'' behavior.

### 2017-07-25 sec-results.tex
Update paper to incorporate review A and review B.
Also update our response to these two reviewers to be more concrete
about the exact changes we have made to the paper and where to find
them.
Add an intuitive explanation of the EM algorithm.
Incorporate new citation from the editor.
Also shuffle around some figures to make the page layout prettier with
the now longer related work and model "derivation" sections.
Grammar fixes from Grammarly analysis.

\section{Experiment Results}
As an analysis tool, the 2L-HMM model provides with us the following two
patterns to characterize student behavior:
\begin{enumerate*}[label=(\arabic*)]
  \item the latent state representations, and
  \item the latent state transitions.
\end{enumerate*}
Thus to evaluate the proposed model, we conduct experiments to
qualitatively analyze both types of patterns discovered from empirical MOOC
log data.

@@

\begin{figure*}
  \centering
  \begin{subfigure}[t]{0.5\textwidth}
    \includegraphics[width=\textwidth]{figures/text-6state/state0.png}
    \caption{An example ``quiz taking'' state.}
    \label{fig:practice-quiz-state}
  \end{subfigure}%
  \begin{subfigure}[t]{0.5\textwidth}
    \includegraphics[width=\textwidth]{figures/text-6state/state1.png}
    \caption{An example ``lecture viewing'' state.}
    \label{fig:lecture-viewing-state}
  \end{subfigure}
  \caption{Two example states found by the 6-state 2L-HMM. (The naming of
  these states reflects our own interpretation.)}
  \label{fig:states}
\end{figure*}

\begin{figure*}
  \centering
  \begin{subfigure}[t]{0.5\textwidth}
    \includegraphics[width=\textwidth]{figures/text-6state/state4.png}
    \caption{A state from \textretrieval{}.}
  \end{subfigure}%
  \begin{subfigure}[t]{0.5\textwidth}
    \includegraphics[width=\textwidth]{figures/sustain-6state/state4.png}
    \caption{A state from \sustain{}.}
  \end{subfigure}
  \caption{Two similar example states found by the 6-state 2L-HMM.}
  \label{fig:cross-course}
\end{figure*}

@@

We also argue that it is important that the latent state representation be
a Markov model rather than just a discrete distribution over actions in
$\mathbf{A}$ (as would be the case for a traditional single-layer HMM). We
can observe why if we take a closer look at each of the two latent state
representations in Figure~\ref{fig:states} and look at their forum
component (bottom right). We can see that the relative probability of the
forum activities is roughly the same between these states, but the
\emph{transitions} are quite different. In
Figure~\ref{fig:practice-quiz-state} we have a relatively low probability
of walking from the ``forum thread view'' action back to the ``forum thread
list'' action (see the bottom rightmost arc; transition probability $p
\approx 0.17$), but in Figure~\ref{fig:lecture-viewing-state} we actually
observe a much stronger link in this direction (transition probability $p
\approx 0.63$). This difference highlights an important distinction between
these two latent states: in the first you are more likely to visit the
forum \emph{looking for a particular post}, where in the second you are
more likely to visit the forum \emph{to browse existing posts}. Thus,
capturing the action transition matrix within a latent state is important
for capturing detailed insights involving bigrams of actions.

We can also use our model for cross-course behavior analysis. In
Figure~\ref{fig:cross-course} we show two latent state representations
learned by a 6-state 2L-HMM, one from \textretrieval{} and one from
\sustain{}. These two states were chosen as they are the most similar
across the two courses. However, if we look at the transitions we can see
some important differences. First, in the state from \textretrieval{}, we
can see that the probability of returning to the course wiki after viewing
a lecture (transition probability $p \approx 0.24$) is considerably lower
than that probability in the state from \sustain{} (transition probability
$p \approx 0.57$). We also notice that the self-loop for staying in a
lecture activity in \textretrieval{} (transition probability $p \approx
0.70$) is significantly higher probability than it is in the state from
\sustain{} (transition probability $p \approx 0.34$).  This gives us some
insight into the lecture viewing behavior of students in these two MOOCs
which can possibly be a reflection of the course's structure (which, as
demonstrated in \citet{Davis:2016:EDM}, can differ substantially across
different MOOCs).  In \textretrieval{}, students are likely to view lecture
videos in succession directly without first visiting the course wiki,
whereas in \sustain{} students are much more likely to first return the
course wiki before watching the next lecture video. This observation would
be lost if we did not consider the transitions between the actions within
the latent states.

@@

We observe similar behavior in Figure~\ref{fig:sustain-state-evolution}
as we increase $K$ when fitting our 2L-HMM on the \sustain{} MOOC.\ Again,
in the transition between $K=2$ and $K=3$ we discover forum behavior
patterns as a latent state. However, in the transition between $K=3$ to
$K=4$ we instead see a refining of that discovered forum state, where
state 3 captures asymmetric transition probabilities between ``forum thread
view'' and ``forum thread list''. These states could be seen as redundant,
in which case a  setting of $K=3$ may be more appropriate for the
\sustain{} dataset.

@@

A unique property of our model is its ability to capture transitions
between the \emph{behavior patterns themselves} that are captured by the
latent states. In Figure~\ref{fig:trans-avg} we show the latent state
transition diagram for a 3-state 2L-HMM fit on \textretrieval{}. We can
immediately observe two things: \begin{enumerate*}[label=(\arabic*)]
  \item each latent state has a very high ``staying'' probability, and
  \item the prevalence of each latent state matches our intuition.
\end{enumerate*}
In particular, we can see that the forum browsing state (state 2) has
relatively lower probability than the other states as we might expect. It
also makes sense that state 0 has rather high probability as this state
likely captures all sequences where a student logged in to the platform and
then did nothing else (likely checking for updates). If we look at state 1
and state 3, their relative probabilities match our intuition as
well. State 1 seems to capture a more engaged browsing session, where there
is non-negligible probability associated with different activities such as
quiz taking and forum browsing and, importantly, these activities have high
probability symmetric edges (so students are taking quizzes one after the
other, or viewing forum threads in succession). By contrast, state 3 seems
to capture a more passive student, with negligible probability mass
associated with forum activity (with low symmetry in the edges). The link
between ``quiz submit'' and ``quiz start'' (indicating quiz repetition) is
also significantly lower than state 1.

\begin{figure*}
  \centering
  \begin{subfigure}[t]{0.33\textwidth}
    \includegraphics[width=\textwidth,trim={0 3cm 0 2cm}]{figures/trans-comp/trans-avg.png}
    \caption{\label{fig:trans-avg}}
  \end{subfigure}%
  \begin{subfigure}[t]{0.33\textwidth}
    \includegraphics[width=\textwidth,trim={0, 3cm 0 2cm}]{figures/trans-comp/trans-perfect.png}
    \caption{\label{fig:trans-perfect}}
  \end{subfigure}%
  \begin{subfigure}[t]{0.33\textwidth}
    \includegraphics[width=\textwidth,trim={0, 3cm 0 2cm}]{figures/trans-comp/trans-low.png}
    \caption{\label{fig:trans-low}}
  \end{subfigure}
  \caption{The latent state transition diagrams for a 4-state 2L-HMM fit to
  \protect\textretrieval{} for all students (a) compared to only ``perfect''
  students (b) and only ``low'' students (c).}
  \label{fig:trans-comp}
\end{figure*}

@@

Returning to Figure~\ref{fig:trans-comp}, we an also see a difference in
the transitions between the latent states. In particular, look at the
transitions in Figure~\ref{fig:trans-perfect} and
Figure~\ref{fig:trans-low} that leave state 2. In the ``perfect'' group,
nearly all of this probability mass is allocated for the self-loop ($p
\approx 0.93$). In the ``low'' group, this self loop is less strong ($p
\approx 0.80$; most easily seen by noting that the edges leaving state 2
are darker than for the ``perfect'' group). We can perform a similar
experiment to above by producing a ranked list of students $\ell \in
\mathbf{L}_q$ by their staying probability for state 2 (that is, given that
a student $\ell$ is already in state 2, how likely are they to remain there
in the next action sequence?). This can be computed as
\begin{equation}
  p_\ell(2, 2) = \frac{\sum_{t=1}^{T-1} \xi_t^{(\mathbf{o}_\ell)}(2,2)}
  {\sum_{i=1}^K \sum_{t=1}^{T-1} \xi_t^{(\mathbf{o}_\ell)}(2,i)}
\end{equation}
where $\xi$ is defined as before and computed using the Baum-Welch
algorithm. The last column of Table~\ref{table:mean-rank} indicates that
this transition feature also correlates with the achievement group and
results in significant differences in mean rank.

>>>>

\section{Experiment Results}
As an analysis tool, the 2L-HMM model provides us with the following two
patterns to characterize student behavior:
\begin{enumerate*}[label=(\arabic*)]
  \item the latent state representations, and
  \item the latent state transitions.
\end{enumerate*}
Thus to evaluate the proposed model, we conduct experiments to
qualitatively analyze both types of patterns discovered from empirical MOOC
log data.

@@

\begin{figure}
  \centering
  \begin{subfigure}[t]{0.5\textwidth}
    \includegraphics[width=\textwidth]{figures/text-6state/state0.png}
    \caption{An example ``quiz taking'' state.}
    \label{fig:practice-quiz-state}
  \end{subfigure}%
  \begin{subfigure}[t]{0.5\textwidth}
    \includegraphics[width=\textwidth]{figures/text-6state/state1.png}
    \caption{An example ``lecture viewing'' state.}
    \label{fig:lecture-viewing-state}
  \end{subfigure}
  \caption{Two example states found by the 6-state 2L-HMM. (The naming of
  these states reflects our own interpretation.)}
  \label{fig:states}
\end{figure}

@@

We also argue that it is important that the latent state representation is
a Markov model rather than just a discrete distribution over actions in
$\mathbf{A}$ (as would be the case for a traditional single-layer HMM). We
can observe why if we take a closer look at each of the two latent state
representations in Figure~\ref{fig:states} and look at their forum
component (bottom right). We can see that the relative probability of the
forum activities is roughly the same between these states, but the
\emph{transitions} are quite different. In
Figure~\ref{fig:practice-quiz-state} we have a relatively low probability
of walking from the ``forum thread view'' action back to the ``forum thread
list'' action (see the bottom rightmost arc; transition probability $p
\approx 0.17$), but in Figure~\ref{fig:lecture-viewing-state} we actually
observe a much stronger link in this direction (transition probability $p
\approx 0.63$). This difference highlights an important distinction between
these two latent states: in the first you are more likely to visit the
forum \emph{looking for a particular post}, where in the second you are
more likely to visit the forum \emph{to browse existing posts}. Thus,
capturing the action transition matrix within a latent state is important
for capturing detailed insights involving bigrams of actions.

\begin{figure}
  \centering
  \begin{subfigure}[t]{0.5\textwidth}
    \includegraphics[width=\textwidth]{figures/text-6state/state4.png}
    \caption{A state from \textretrieval{}.}
  \end{subfigure}%
  \begin{subfigure}[t]{0.5\textwidth}
    \includegraphics[width=\textwidth]{figures/sustain-6state/state4.png}
    \caption{A state from \sustain{}.}
  \end{subfigure}
  \caption{Two similar example states found by the 6-state 2L-HMM.}
  \label{fig:cross-course}
\end{figure}

We can also use our model for cross-course behavior analysis. In
Figure~\ref{fig:cross-course} we show two latent state representations
learned by a 6-state 2L-HMM, one from \textretrieval{} and one from
\sustain{}. These two states were chosen as they are the most similar
between the two courses. However, if we look at the transitions we can see
some important differences. First, in the state from \textretrieval{}, we
can see that the probability of returning to the course wiki after viewing
a lecture (transition probability $p \approx 0.24$) is considerably lower
than that probability in the state from \sustain{} (transition probability
$p \approx 0.57$). We also notice that the self-loop for staying in a
lecture activity in \textretrieval{} (transition probability $p \approx
0.70$) is significantly higher probability than it is in the state from
\sustain{} (transition probability $p \approx 0.34$).  This gives us some
insight into the lecture viewing behavior of students in these two MOOCs
which can possibly be a reflection of the course's structure (which, as
demonstrated in \citet{Davis:2016:EDM}, can differ substantially across
different MOOCs).  In \textretrieval{}, students are likely to view lecture
videos in succession directly without first visiting the course wiki,
whereas in \sustain{} students are much more likely to first return the
course wiki before watching the next lecture video. This observation would
be lost if we did not consider the transitions between the actions within
the latent states.

@@

We observe similar behavior in Figure~\ref{fig:sustain-state-evolution}
as we increase $K$ when fitting our 2L-HMM on the \sustain{} MOOC.\ Again,
in the transition between $K=2$ and $K=3$, we discover forum behavior
patterns as a latent state. However, in the transition between $K=3$ to
$K=4$, we instead see a refining of that discovered forum state, where
state 3 captures asymmetric transition probabilities between ``forum thread
view'' and ``forum thread list''. These states could be seen as redundant,
in which case a  setting of $K=3$ may be more appropriate for the
\sustain{} dataset.

@@

A unique property of our model is its ability to capture transitions
between the \emph{behavior patterns themselves} that are captured by the
latent states. In Figure~\ref{fig:trans-avg} we show the latent state
transition diagram for a 4-state 2L-HMM fit on \textretrieval{}. We can
immediately observe two things: \begin{enumerate*}[label=(\arabic*)]
  \item each latent state has a very high ``staying'' probability, and
  \item the prevalence of each latent state matches our intuition.
\end{enumerate*}
In particular, we can see that the forum browsing state (state 2) has
relatively lower probability than the other states as we might expect. It
also makes sense that state 0 has a rather high probability as this state
likely captures all sequences where a student logged in to the platform and
then did nothing else (likely checking for updates). If we look at state 1
and state 3, their relative probabilities match our intuition as
well. State 1 seems to capture a more engaged browsing session, where there
is non-negligible probability associated with different activities such as
quiz taking and forum browsing and, importantly, these activities have high
probability symmetric edges (so students are taking quizzes one after the
other, or viewing forum threads in succession). By contrast, state 3 seems
to capture a more passive student, with negligible probability mass
associated with forum activity (with low symmetry in the edges). The link
between ``quiz submit'' and ``quiz start'' (indicating quiz repetition) is
also significantly lower than state 1.

\begin{figure}
  \centering
  \begin{subfigure}[t]{0.33\textwidth}
    \includegraphics[width=\textwidth,trim={0 3cm 0 2cm}]{figures/trans-comp/trans-avg.png}
    \caption{\label{fig:trans-avg}}
  \end{subfigure}%
  \begin{subfigure}[t]{0.33\textwidth}
    \includegraphics[width=\textwidth,trim={0, 3cm 0 2cm}]{figures/trans-comp/trans-perfect.png}
    \caption{\label{fig:trans-perfect}}
  \end{subfigure}%
  \begin{subfigure}[t]{0.33\textwidth}
    \includegraphics[width=\textwidth,trim={0, 3cm 0 2cm}]{figures/trans-comp/trans-low.png}
    \caption{\label{fig:trans-low}}
  \end{subfigure}
  \caption{The latent state transition diagrams for a 4-state 2L-HMM fit to
  \protect\textretrieval{} for all students (a) compared to only ``perfect''
  students (b) and only ``low'' students (c).}
  \label{fig:trans-comp}
\end{figure}

@@

Returning to Figure~\ref{fig:trans-comp}, we can also see a difference in
the transitions between the latent states. In particular, look at the
transitions in Figure~\ref{fig:trans-perfect} and
Figure~\ref{fig:trans-low} that leave state 2. In the ``perfect'' group,
nearly all of this probability mass is allocated for the self-loop ($p
\approx 0.93$). In the ``low'' group, this self-loop is less strong ($p
\approx 0.80$; most easily seen by noting that the edges leaving state 2
are darker than for the ``perfect'' group). We can perform a similar
experiment to above by producing a ranked list of students $\ell \in
\mathbf{L}_q$ by their staying probability for state 2 (that is, given that
a student $\ell$ is already in state 2, how likely are they to remain there
in the next action sequence?). This can be computed as
\begin{equation}
  p_\ell(2, 2) = \frac{\sum_{t=1}^{T-1} \xi_t^{(\mathbf{o}_\ell)}(2,2)}
  {\sum_{i=1}^K \sum_{t=1}^{T-1} \xi_t^{(\mathbf{o}_\ell)}(2,i)}
\end{equation}
where $\xi$ is defined as before and computed using the Baum-Welch
algorithm. The last column of Table~\ref{table:mean-rank} indicates that
this transition feature also correlates with the achievement group and
results in significant differences in mean rank.
